A practical guide to Digital Micro-mirror Devices (DMDs) for wavefront shaping

### 2023-11-15 paper/main.tex
update text

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\usepackage{graphicx}
\usepackage{subcaption}

@@

\usepackage[most]{tcolorbox}

@@

\newtcolorbox{tldr}[1][]{%
enhanced jigsaw,
%oversize,
colback=light_orange,
boxrule=0pt,
overlay unbroken and first ={%
\draw[line width=0.2pt,double=bg,draw=bg!70!black,
    double distance=1pt,] (frame.north west) -- (frame.north east);
\draw[line width=0.2pt,double=bg,draw=bg!70!black,
    double distance=1pt,] (frame.south west) -- (frame.south east);},
breakable,
arc=0pt,outer arc=0pt,
#1}%

@@

@@

Since the advent of adaptive optics, various technologies have been employed 
to modulate the amplitude and/or phase of light. 
Early adaptive optics devices, utilized in fields like microscopy and astronomy, 
offer rapid modulation capable of compensating for the aberrations of optical systems 
in real-time. 
However, these devices are constrained by a limited number of actuators, 
restricting their utility in complex media where a large number of degrees of freedom is essential. 
Liquid Crystal Spatial Light Modulators (LC-SLMs), 
which allow for the control of light phase across typically more than a million pixels, 
have emerged as powerful tools for wavefront shaping in complex media 
since the seminal work of A. Most and I. Vellekoop in the mid-2000s~\cite{Vellekoop2007focusing}. 
Nonetheless, LC-SLMs are hampered by their slow response time, 
permitting only a modulation speed ranging from a few Hz to about 100 Hz. 
Digital Micromirror Devices (DMDs) have emerged as a technology bridging the gap 
between these two types of systems; 
they offer a large number of pixels (similar to LC-SLMs) and fast modulation speeds (typically up to several tens of kHz). 
However, DMDs are restricted to binary amplitude modulation and are not optimized for coherent light applications. 
Utilizing DMDs for coherent control of light in complex media is therefore non-trivial 
and necessitates specific adaptations for efficient use. 
This tutorial aims to provide a comprehensive guide 
for selecting and employing DMDs in wavefront shaping applications within complex media.\\

To comprehend both the capabilities and limitations
of DMD technology for coherent wavefront shaping,
it is crucial to understand the device's operating principles
and its original design intentions.
Investigated and developed by Texas Instruments since the late 1970s,
DMDs gained prominence in the 1990s for video projection applications
under the commercial name of Digital Light Processing (DLP)~\cite{Dudley2003emerging}.
The technology enables high-resolution, high-speed, and high-contrast-ratio modulation of light.\\
DMDs operate by toggling the state of small mirrors between two distinct angles, denoted as $\pm \theta_\text{DMD}$.
The device is originally engineered for amplitude modulation in video projection applications.
In this configuration, one mirror angle directs light into the projection lens,
while the alternate angle results in the light path being blocked (see Fig~\ref{fig:combined_pixel}).
Given that projectors utilize incoherent light and that the DMD plane is optically conjugated with the projection screen,
aberrations within the DMD plane are generally not problematic.
Similarly, phase fluctuations induced by temperature variations, 
as well as minor vibrations from the cooling hardware, are inconsequential in this context.
The DMD is designed to produce binary on/off modulation,
which is then leveraged to generate grayscale images via pulse-width modulation.
Color modulation is accomplished through the use of a color wheel in conjunction with a bright white light source.\\

Third-party companies have developped kits for research applications,
providing a DMD with a control board and a software interface. 
In particular, Vialux devices~\cite{}, provides a FPGA board that enable high speed modulation 
by allowing frames to be stored in the device memory. 
However, sandard Texas Instrument video projector evaluation modules can be converted into a wavefront shaping device~\cite{Cox2021converting}, 
albeit with a lower modulation speed.
Such systems can then be converted into phase or complex field modulators. 
It typically relies on encoding the optical phase into 
the spatial displacement of binaray spatial fringes displayed on the DMD 
and then filtering the high spatial frequencies in the Fourier plane~\cite{Lee1979binary}.
It allows multiple level complex modulation at the expense of a reduced spatial resolution.
The implementation and the performance of such a system are studied in a separate tutorial~\cite{RODRIGO} 
and will not be detailed here.
In the rest of the paper, we will assume that the DMD is used for complex modulation using such an approach.
While the present tutorial is not specific to any particular device, 
the software section will focus on the Vialux control board.
In a first part....\\

@@

% \cite{JacksonDMD}

@@

One important difference between liquid crystal modulators and DMDs 
concerns the geometry of the pixel surface. 
This leads to diffraction effects that can be detrimental 
to the modulation and efficiency of the system.
It strongly depends on the wavelength of the illumination, 
the pixel pitch, and the incident and outgoing angles. 
Thus, in addition to choosing an appropriate anti-reflection coating, 
it is important to check that the pixel pitch is well-suited for the configuration used.
Texas Instruments chips come with different possible pixel pitch $d$ 
ranging roughly from $5$ to $\sim25$\textmu m\cite{TI}.\\

Let's consider a 1D system,
as illustrated in Fig.~\ref{fig:grating_geom},  
% For the sake of simplicity and 
to qualitatively understand the problem.
% we place ourselves in the small angle approximation. 
This corresponds to incident and outgoing angles close to normal to the modulator surface 
and to a small tilt angle of the mirror. 
% We will give latter the final formulas in the 2D case.
We first assume that all the pixels are in the same state 
and are illuminated by a plane wave from the far field.
A pixelated modulator then acts exactly as a grating, 
with a period $d$ equal to the pixel pitch.
Modulators have a filling fraction limited by the hardware, 
typically around 90\%.
This corresponds to having an effective size of active pixels of $d' < d$.

In general, a grating leads to the observation of diffraction orders  
with varying intensities 
and angles $\theta_p$ given by the grating equation
$\text{sin}(\theta_p) = p\lambda/d = p \text{sin}(\theta_D)$,
where $\lambda$ is the wavelength of the light 
and $\theta_D$ the angle of the first order of diffraction.
The intensity of the diffraction orders depends on the size and the response of one pixel.
We see that we can decouple the effect of the periodicity of the grating, 
that acts on the angles of the orders of diffraction, 
and the response of one pixel, 
that acts on the envelope of the angular response.

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{images/grating_geom.pdf}
  \label{fig:flat_grating}

  \caption{
  \textbf{1D grating geometry.}
 ...
  }
  \label{fig:grating_geom}
  \end{figure}

For a normal incidence 
in the case of a LC-SLM, the response of one pixel is 
assumed constant over the pixel size. 
For a blazed grating, we also have a linear phase slope on each pixel due to the tilt $\theta_B$ of the pixels.
For an an arbitrary incidence angle $\alpha$, we also have a global phase slope 
that has the effect of trivially shifting the angular diffraction by an angle $\alpha$.\\

In essence, the incident angle $\alpha$ has the effect to shift the whole angular diffraction pattern 
with respect to the normal incidence case, 
while the blazed angle $\theta_B$, the tilt angle of the mirror in the DMD, 
has the effect to shift only the envelope by $2\theta_B$.
When the filling fraction is close to 100\%, 
i.e. $d\approx d'$,
the enveloppe for a flat grating is maximal 
at $\theta_0=-\alpha$ corresponding to the angle of the zero-th order,
and close to zero for the other orders.
We then observe a single diffraction order, 
which corresponds to the ideal case.
Adding a blazed angle $\theta_B$ leads to a shift of the envelope, 
the position of the maximum 
$\theta_0=2\theta_B-\alpha$ 
may not correspond anymore to a diffraction order.\\

\begin{optional}
  If we neglect the finite size of the device and illumination 
  and place ourselves in the small angle approximation, 
  we can write the field reflected off the device 
  for a plane wave illumination
  in the two systems as:

  \begin{equation}
  \begin{aligned}
    R_\text{flat}(x) &\propto \left[\Pi\left(x/d'\right) \otimes_x \sum_k \delta(x-k d)\right]e^{-j\frac{2\pi}{\lambda}\text{sin}(\alpha) x} \\
    R_\text{blazed}(x) &\propto 
        \underbrace{
          \left(
            \underbrace{\Pi\left(x/d'\right)}_\text{pixel size}
              \times 
              \underbrace{e^{j\frac{2\pi}{\lambda}2\text{sin}(\theta_B-\alpha) x}}_{\substack{\text{blazed angle} \\ \text{+ angle of incidence}}}
              %\text{blazed angle + angle of incidence}
          \right) 
        }_\text{pixel response}
        \otimes_x 
        \left[
        \underbrace{\sum_k \delta(x-k d) }_\text{periodicitiy}
        .
        \underbrace{
        e^{-j\frac{2\pi}{\lambda}\text{sin}(\alpha) kd}
      }_\text{angle of incidence}
      \right]
      \, ,
  \end{aligned}
  \label{eq:grating_response}
  \end{equation}

@@

  The intensity as a function of the angle in the far-field 
  is given, up to a homotetic transformation, 
  by the absolute value squared of 
  Fourier transform of Eq.~\ref{eq:grating_response}
  and reads:

  \begin{equation}
    \begin{aligned}
      I_\text{flat}(\theta) \propto 
        \sum_p \delta(\text{sin}(\theta)+\text{sin}(\alpha)-p\,\theta_D) 
          \times 
        \text{sinc}^2\left( (\text{sin}(\theta)+\text{sin}(\alpha)) \frac{\lambda}{d'}\right)\\
        I_\text{blazed}(\theta) \propto 
        \underbrace{
          \sum_p \delta(\text{sin}(\theta)+\text{sin}(\alpha)-p\,\theta_D) 
        }_\text{orders of diffraction}
        \times 
        \underbrace{
          \text{sinc}^2\left( (\text{sin}(\theta)+\text{sin}(2\theta_B-\alpha)) \frac{\lambda}{d'}\right)
         }_\text{envelope} \, .
    \end{aligned}
  \end{equation}

\end{optional}
A more accurate computation of the far field can be found  in ~\cite{Wang2023diffraction}.
We observe that the envelope (right hand term) is maximal for $\text{sin}(\theta_{max}) = \text{sin}(\alpha-2\theta_B)$ 
while the effect of the periodicity (left hand term) is maximal for $\text{sin}(\theta_p)+ \text{sin}(\alpha) = p \lambda/d$. 
with $p$ an integer value.
We represent in Fig.~\ref{fig:gratings} the angular response of a flat grating and a blazed grating 
for a 1D filling fraction of 95\% (correponding to a 2D filling fraction of $\approx 90$\%).
For a flat grating, only the zero-th order has a significant and comparable intensity. 
For a blazed grating, in this example close to the worst case scenario, 
two diffraction orders have a significant intensity, 
and other orders also have a non-negligible contributions. 
% This implies that by selecting one order, we can loose more that 50\% of the light. 
% In a 2D system, we can then loose more that 75\% of the light.\\ 
The ideal case, i.e. when the maximum of the envelope corresponds to a diffraction order,
leading a single diffraction order containing most of the energy,
is obtained when the blazed grating equation is satistied~\cite{Casini2014on}:

@@

% \begin{equation}
%     a = 
%   \left\{
%     \begin{aligned}
%         a\\b
% %         x + y   &= 7 \\
% %         5x - 2y &= -7
%     \end{aligned}
%   \right\}
% \end{equation}

@@

To analyze more precisely the effect of diffraction in a DMD, 
we need to consider the 2D surface of the modulator.
Let's define a cartesian coordinate system in the plane of the DMD
with axis $x$ and $y$ aligned with the pixel sides 
(see Fig.~\ref{fig:2d_geom}a.). 
Pixels are repeated along those directions. 
A technical difficulty is that the axis of rotation of the pixels 
is the diagonal of the pixels, 
hence rotated by 45 degrees with respect to the $x$ and $y$. 
For the alignment and the maniputlation of the optical setup
it is more convenient to work in incident and outgoing beams 
with optical axis contained in an horizontal plane. 
A simple and common solution consists in roatating the chip by 45 degrees
with respect to the horizontal plane 
such that the axis of rotation of the pixels is vertical.  
This configuration is illustrated in Fig.~\ref{fig:2d_geom}.

\begin{figure}
  \centering
  \includegraphics[width = 0.8\textwidth]{images/dmd_45.pdf}
  \caption{
  \textbf{2D grating geometry.}
 ...
  }
  \label{fig:2d_geom}
\end{figure}

The 2D array can be seen as 2 orthogonal 1D gratings in the $x$ and $y$ directions.
The incident angle $\alpha'$. 
Because of the 45 degrees angle between 
axis of rotation of the mirrors and axis, 
both 1D gratings have the same properties. 
A more complete computation of the 2D system can be found in~\cite{Scholes2019structured}.
To apply Eq.~\ref{eq:blazed_eq}, 
we first need to project the 
the different angles of the problem in the incident planes 
of the two 1D gratings.
It gives 
$\alpha = \text{arctan}\left(\text{tan}(\alpha')/\sqrt{2}\right)$, 
$\beta = \text{arctan}\left(\text{tan}(\beta')/\sqrt{2}\right)$, 
and $\theta_B = \text{arctan}\left(\text{tan}(\theta_\text{DMD})/\sqrt{2}\right)$, 
$theta_\text{DMD}$ being the actual rotation angle of the mirrors with respect to the diagonal of the pixels.
We can quantify how close we are to the ideal case, 
i.e. when satifying the blazed equation~\ref{eq:blazed_eq}, 
by defining a {\em blazed number} $\mu$ as~\cite{WFSnet_diffraction}:

\begin{equation}
  \mu =  
  \left|
    \lfloor 2 \frac{d}{\lambda} 
      \left[
        \text{sin}(\alpha)  +\text{sin}(\beta)
      \right] 
    \rfloor
    \mod{2} -1 
  \right| \, ,
\end{equation}

with $\lfloor . \rfloor$ representing the integer part, 
and $\mod{2}$ the modulo 2 operation.
$\mu$ is maximal when the blazing equation is satisfield, 
i.e. when one order of diffraction contains most of the energy, 
and minimal when we are in the worst case scenario, 
i.e. when four orders of diffraction have a significant and equal intensity.\\

To illustrate the effect, we simulate a DMD in Python 
(see tutorial and code in~\cite{WFSnet_diffraction})
with a two pixel pitches of 
$d=7.6$\textmu m and $d=10.8$\textmu m,
for a coherent excitation at $\lambda=633$nm. 
We show in Fig.~\ref{fig:mu76} the blazed number $\mu$
as a function of the incident angle $\alpha'$ 
as well as the far field diffraction pattern for different two incident angles.
While one can change the efficiency of diffraction by tuning 
the incindent angle, its effect is somewhat limited 
withing a reasonable angular range compatible with experimental constraints 
(i.e. for angles far from $\pm 90^\circ$).
Far field pattern are centered around the maximum of the envelope 
$\theta_\text{max} = 2\theta_B - \alpha$ 
(marked by a yellow cross).
We see that for small values of $\alpha$, 
the pixel pitch of $d=10.8$\textmu m leads to a blazed number $\mu$ 
close $1$. 
It corresponds in the far field to having one bright order of diffraction  
close to the maximum of the envelope.

\begin{figure}
  \centering
  \begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = \textwidth]{images/mu_76.pdf}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = \textwidth]{images/mu_108.pdf}
  \end{subfigure}
  \caption{
  \textbf{Blazed number and far field diffraction pattern for...}
 ...
  }
  \label{fig:mu76}
\end{figure}

@@

\section{Control Vialux DMDs with Python}

\cite{hofling2004alp}

\cite{popoff2016alp4lib}

>>>>

\documentclass[12pt]{iopart}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

% !TEX root =main.tex

\usepackage{graphicx}
\usepackage{subcaption}

@@

\usepackage[most, breakable]{tcolorbox}
\usepackage{hyperref}

\hypersetup{
    breaklinks = true,
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=blue,
    citecolor=red
}

@@

\newtcolorbox{tldr}[1][]{%
enhanced jigsaw,
%oversize,
colback=light_orange,
boxrule=0pt,
overlay unbroken and first ={%
\draw[line width=0.2pt,double=bg,draw=bg!70!black,
    double distance=1pt,] (frame.north west) -- (frame.north east);
\draw[line width=0.2pt,double=bg,draw=bg!70!black,
    double distance=1pt,] (frame.south west) -- (frame.south east);},
arc=0pt,outer arc=0pt,
#1}%

@@

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}

@@

Since the advent of adaptive optics, various technologies have been employed
to modulate the amplitude and/or phase of light.
Early adaptive optics devices, utilized in fields like microscopy and astronomy,
offer rapid modulation capable of compensating for the aberrations of optical systems
in real-time.
However, these devices are constrained by a limited number of actuators,
restricting their utility in complex media where a large number of degrees of freedom is essential.
Liquid Crystal Spatial Light Modulators (LC-SLMs),
which allow for the control of light phase across typically more than a million pixels,
have emerged as powerful tools for wavefront shaping in complex media
since the seminal work of A. Most and I. Vellekoop in the mid-2000s~\cite{Vellekoop2007focusing}.
Nonetheless, LC-SLMs are hampered by their slow response time,
permitting only a modulation speed ranging from a few Hz to about 100 Hz.
Digital Micromirror Devices (DMDs) have emerged as a technology bridging the gap
between these two types of systems;
they offer a large number of pixels (similar to LC-SLMs) and fast modulation speeds (typically up to several tens of kHz).
However, DMDs are restricted to binary amplitude modulation and are not optimized for coherent light applications.
Utilizing DMDs for coherent control of light in complex media is therefore non-trivial
and necessitates specific adaptations for efficient use.
While other articles exist describing the various aspects of DMDs~\cite{Park2015properties, Cox2021converting, Wang2023diffraction},
this tutorial aims to provide a guide for easily setting up a DMD for wavefront shaping applications in complex media.
In particular, we provide characterization and validation procedures that requires minimal changes compared to typical wavefront shaping setups.\\

REVIEW en bio imaging~\cite{Zhuang2020}

To comprehend both the capabilities and limitations
of DMD technology for coherent wavefront shaping,
it is crucial to understand the device's operating principles
and its original design intentions.
Investigated and developed by Texas Instruments since the late 1970s,
DMDs gained prominence in the 1990s for video projection applications
under the commercial name of Digital Light Processing (DLP)~\cite{Dudley2003emerging}.
The technology enables high-resolution, high-speed, and high-contrast-ratio modulation of light.\\
DMDs operate by toggling the state of small mirrors between two distinct angles, denoted as $\pm \theta_\text{DMD}$.
The device is originally engineered for amplitude modulation in video projection applications.
In this configuration, one mirror angle directs light into the projection lens,
while the alternate angle results in the light path being blocked (see Fig.~\ref{fig:combined_pixel}).
Given that projectors utilize incoherent light and that the DMD plane is optically conjugated with the projection screen,
aberrations within the DMD plane are generally not problematic.
Similarly, phase fluctuations induced by temperature variations,
as well as minor vibrations from the cooling hardware, are inconsequential in this context.
The DMD is designed to produce binary on/off modulation,
which is then leveraged to generate grayscale images via pulse-width modulation.
Color modulation is accomplished through the use of a color wheel in conjunction with a bright white light source.\\

Third-party companies have developed kits tailored for research applications,
which include a DMD, a control board, and a software interface.
Specifically, Vialux devices~\cite{vialux} offer an FPGA board that enables high-speed modulation
by allowing frames to be stored in the device's memory~\cite{hofling2004alp}.
However, standard Texas Instruments video projector evaluation modules can also be repurposed into wavefront shaping devices~\cite{Cox2021converting},
though at a compromised modulation speed.
These systems can further be converted into phase or complex field modulators.
This is typically achieved by encoding the optical phase into
the spatial displacement of binary spatial fringes displayed on the DMD,
followed by filtering the high spatial frequencies in the Fourier plane~\cite{lee1979binary}.
Such a configuration permits multi-level complex modulation but sacrifices spatial resolution.
The implementation and performance of such systems are explored in a separate tutorial~\cite{Gutierrez2024DMD}
and will not be elaborated upon here.
For the remainder of this paper, it will be assumed that the DMD is used for complex modulation via such a method.
While the present tutorial is not constrained to any specific device,
the software section will predominantly focus on the Vialux control board.
In the first part....\\

@@

@@

A significant distinction between liquid crystal modulators and DMDs
lies in the geometry of the pixel surface.
This difference gives rise to diffraction effects that can adversely affect
both the modulation quality and system efficiency.
The impact of these diffraction effects is highly contingent on several factors:
the wavelength of the illumination, the pixel pitch, and both the incident and outgoing angles.
Therefore, alongside selecting an appropriate anti-reflection coating,
it is crucial to ensure that the pixel pitch is compatible with the specific configuration being used.
Texas Instruments offers chips with a variety of pixel pitches $d$,
ranging approximately from $5$ to $\sim25$ \textmu m~\cite{TI}.\\

To achieve a qualitative understanding of this issue,
we consider a 1D array of pixels as illustrated in Fig.~\ref{fig:grating_geom}.
% For simplicity, we will focus on scenarios where both the incident and outgoing angles 
% are nearly perpendicular to the modulator surface, and the mirror tilt angle is minimal.
Initially, let's assume that all pixels are in the same state and are illuminated by a plane wave originating from the far field.
Under these conditions, the pixelated modulator essentially functions as a grating,
with a period $d$ that is equivalent to the pixel pitch.
It is important to underscore that these modulators possess a hardware-limited fill factor,
typically around 90\%.
This translates to an effective active pixel size of $d' < d$.\\

In general, a grating gives rise to various diffraction orders
with differing intensities
and angles $\theta_p$, as dictated by the grating equation:
$
  \text{sin}(\theta_p) = p\lambda/d = p \, \text{sin}(\theta_D),
$
where $\lambda$ is the wavelength of the light,
and $\theta_D$ is the angle of the first-order diffraction.
The intensity of the individual diffraction orders is influenced by both the size and the response of a single pixel.
Importantly, we can decouple the effects of the grating's periodicity,
which influences the angles of the diffraction orders,
from the effects of the response of a single pixel,
which shapes the envelope of the angular response.\\

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{images/grating_geom.pdf}
  \caption{
    \textbf{1D grating geometry.}
    Schematic representation of the geometry of two types of modulators:
    (a) the liquid crystal modulator, equivalent to a flat grating,
    and (b) the DMD geometry, equivalent to a blazed grating.
    $\alpha$ denotes the incident angle relative to the normal of the array plane,
    $\theta_0$ refers to the angle of the zeroth diffraction order,
    and $\theta_B$ is the tilt angle of the mirrors.\\
  }
  \label{fig:grating_geom}
\end{figure}

For a case of normal incidence with an LC-SLM (Liquid Crystal Spatial Light Modulator),
we can assume that the response of a single pixel is uniform its surface.
Similarly, in the scenario of a blazed grating,
such as a DMD, a linear phase slope is present on each pixel.
This is due to the tilt angle $\theta_B$ of the mirrors.
For an arbitrarily selected incidence angle $\alpha$,
a global phase slope is introduced.
This results in a trivial shift of the angular diffraction pattern by an angle $\alpha$.
In essence, the incident angle $\alpha$ serves to shift the entire angular diffraction pattern
relative to the case of normal incidence,
while the blazed angle $\theta_B$—the tilt angle of the mirror in the DMD—
only shifts the envelope by an angle of $2\theta_B$.
Whenever the fill factor approaches 100\%,
i.e. when $d \approx d'$,
the envelope for a flat grating achieves its maximum at $\theta_0 = -\alpha$;
this corresponds to the angle of the zeroth-order diffraction,
which nears zero for all other orders.
In these specific conditions, a singular diffraction order is visually perceived, corresponding to the optimum scenario.
The addition of a blazed angle $\theta_B$ results in both
a shift in the envelope and in the position of its maximum,
now indicated by $\theta_0 = 2\theta_B - \alpha$.
In the general case,
this position may not align with a diffraction order anymore~\cite{Park2015properties}.

\begin{optional}
  Assuming the effect of the device's finite size and illumination to be negligible,
  and situating ourselves within the context of the small-angle approximation,
  we can represent the field reflected from the device
  under the influence of plane wave illumination
  across two systems as follows:

  \begin{equation}
    \begin{aligned}
      R_\text{flat}(x)   & \propto \left[\Pi\left(x/d'\right) \otimes_x \sum_k \delta(x-k d)\right]e^{-j\frac{2\pi}{\lambda}\text{sin}(\alpha) x} \\
      R_\text{blazed}(x) & \propto
      \underbrace{
        \left(
        \underbrace{\Pi\left(x/d'\right)}_\text{pixel size}
        \times
      \underbrace{e^{j\frac{2\pi}{\lambda}2\text{sin}(\theta_B-\alpha) x}}_{\substack{\text{blazed angle}                                         \\ \text{+ angle of incidence}}}
        %\text{blazed angle + angle of incidence}
        \right)
      }_\text{pixel response}
      \otimes_x
      \left[
        \underbrace{\sum_k \delta(x-k d) }_\text{periodicitiy}
        .
        \underbrace{
          e^{-j\frac{2\pi}{\lambda}\text{sin}(\alpha) kd}
        }_\text{angle of incidence}
        \right]
      \, ,
    \end{aligned}
    \label{eq:grating_response}
  \end{equation}

@@

  The intensity as a function of the angle in the far-field
  is given, up to a homotetic transformation,
  by the absolute value squared of

  The Fourier transform of Eq.~\ref{eq:grating_response} can be written as:

  \begin{equation}
    \begin{aligned}
      I_\text{flat}(\theta) \propto
      \sum_p \delta(\text{sin}(\theta)+\text{sin}(\alpha))
      \times
      \text{sinc}^2\left( \left[\text{sin}(\theta)+\text{sin}(\alpha)\right] \frac{\lambda}{d'}\right) \\
      I_\text{blazed}(\theta) \propto
      \underbrace{
        \sum_p \delta(\text{sin}(\theta)+\text{sin}(\alpha)-p\,\theta_D)
      }_\text{orders of diffraction}
      \times
      \underbrace{
        \text{sinc}^2\left( \left[\text{sin}(\theta)+\text{sin}(\alpha)\right] \frac{\lambda}{d'}\right)
      }_\text{envelope} \, .
    \end{aligned}
  \end{equation}

\end{optional}
A more accurate computation of the far field can be found  in ~\cite{Wang2023diffraction}.
We observe that the envelope (right hand term) is maximal for $\text{sin}(\theta_{max}) = \text{sin}(\alpha-2\theta_B)$
while the effect of the periodicity (left hand term) is maximal for $\text{sin}(\theta_p)+ \text{sin}(\alpha) = p \lambda/d$.
with $p$ an integer value.
We represent in Fig.~\ref{fig:gratings} the angular response of a flat grating and a blazed grating
for a 1D filling fraction of 95\% (correponding to a 2D filling fraction of $\approx 90$\%).
For a flat grating, only the zero-th order has a significant and comparable intensity.
For a blazed grating, in this example close to the worst case scenario,
two diffraction orders have a significant intensity,
and other orders also have a non-negligible contributions.
In the optimal scenario, where the peak of the envelope corresponds to a diffraction order,
it results in a single diffraction order carrying the majority of the energy.
This state is achieved when the conditions of the blazed grating equation are fulfilled~\cite{Casini2014on}:

@@

@@

To analyze more precisely the effect of diffraction in a DMD,
one needs to consider the 2D surface of the modulator.
We can establish a Cartesian coordinate system on the plane of the DMD,
with axes $x$ and $y$ aligned with the pixel sides
(refer to Fig.~\ref{fig:2d_geom}a).
Pixels are uniformly repeated along these axes.
However, a technical challenge arises
in that the axis of rotation of the pixels
aligns with the pixel diagonals,
resulting in a rotation by 45 degrees with respect to the $x$ and $y$ axes.
For the convenience of alignment and manipulation of the optical setup,
it is preferable to work with the incident and outgoing beams
which have the optical axis contained in the horizontal plane.
A straightforward and prevalent solution is to rotate the chip by 45 degrees
relative to the horizontal plane,
which aligns the pixel’s axis of rotation to be vertical.
This configuration is depicted in Fig.~\ref{fig:2d_geom}b.

\begin{figure}
  \centering
  \includegraphics[width = 0.9\textwidth]{images/dmd_45.pdf}
  \caption{
    \textbf{2D grating geometry.}
    (a) Schematic representation of the geometry of the DMD.
    (b) Photograph of the DMD chip oriented so that the rotation axis of the pixels
    is aligned vertically.
  }
  \label{fig:2d_geom}
\end{figure}

The 2D array can be seen as 2 orthogonal 1D gratings in the $x$ and $y$ directions,
both having the same properties.
A more comprehensive depiction of the 2D system can be found in~\cite{Scholes2019structured}.
In order to utilize Eq.(\ref{eq:blazed_eq}),
one first needs projecting
the different angles of the problem onto the incident planes
of the two 1D gratings.
This yields
$\alpha = \text{arctan}\left(\text{tan}(\alpha')/\sqrt{2}\right)$,
$\beta = \text{arctan}\left(\text{tan}(\beta')/\sqrt{2}\right)$,
and $\theta_B = \text{arctan}\left(\text{tan}(\theta_\text{DMD})/\sqrt{2}\right)$,
where $\theta_\text{DMD}$ represents the true rotation angle of the mirrors relative to the diagonal of the pixels,
and $\alpha'$ and $\beta'$ denote the incident and outgoing angles in the horizontal plane.
We can quantify how close we are to the ideal case,
i.e., when satifying the blazed equation outlined by Eq.(\ref{eq:blazed_eq}),
by defining a {\em blazed number} $\mu$ as introduced in~\cite{WFSnet_diffraction}:

\begin{equation}
  \mu =
  \left|
  \lfloor 2 \frac{d}{\lambda}
  \left[
    \text{sin}(\alpha)  +\text{sin}(\beta)
    \right]
  \rfloor
  \mod{2} -1
  \right| \, ,
  \label{eq:blazed_number}
\end{equation}

with $\lfloor . \rfloor$ representing the integer part,
and $\mod{2}$ the modulo 2 operation.
$\mu$ is maximal and equals $1$ when the blazing equation is satisfield,
i.e. when one order of diffraction contains most of the energy,
and minimal when we are in the worst case scenario,
i.e. when four orders of diffraction have a significant and equal intensity.\\

To demonstrate the effect, we conduct a simulation of a DMD using Python
(refer to tutorial and code in~\cite{WFSnet_diffraction})
with two pixel pitches of
$d=7.6$\textmu m and $d=10.8$\textmu m,
under a coherent excitation at $\lambda=633$nm.
Fig.~\ref{fig:mu76} shows the estimated blazed number $\mu$
as a function of the angle of incidence $\alpha'$,
along with the far field diffraction pattern for two distinct incident angles.
It should be noted that the efficiency of diffraction can be altered by adjusting
the angle of incidence, yet its impact is relatively confined
within an acceptable angular range that aligns with experimental limitations
(i.e. for angles far from $\pm 90^\circ$).
Far field patterns are centered around the maximum of the envelope
$\theta_\text{max} = 2\theta_B - \alpha$
(marked by a yellow cross).
We see that for small values of $\alpha$,
the pixel pitch of $d=10.8$\textmu m leads to a blazed number $\mu$
close $1$.
It corresponds in the far field to having one bright order of diffraction
close to the maximum of the envelope.

\begin{figure}
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{images/mu_76.pdf}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{images/mu_108.pdf}
  \end{subfigure}
  \caption{
    \textbf{Blazed number and far field diffraction patterns, extreme cases.}
    Blazing number $\mu$ (Eq.(\ref{eq:blazed_number})) as a function of the incident angle $\alpha'$  (top)
    for a pixel pitch of $d=7.6$\textmu m (left)
    and $d=10.8$\textmu m (right).
    Corrseponding far field diffraction pattern (bottom)
    for two incident angles $\alpha' = 0^\circ$ and $\alpha' = 25^\circ$.
    The yellow cross indicates the maximum of the envelope $\theta_\text{max} = 2\theta_B - \alpha$.
  }
  \label{fig:mu76}
\end{figure}

@@

% \section{Control Vialux DMDs with Python}

% \cite{hofling2004alp}

% \cite{popoff2016alp4lib}
